Combining information from photometric and spectroscopic data:\\ propogating uncertainties in spectrophotometric calibration

### 2015-04-24 tex/ms.tex
title change and section reorg, some changes to the mock parameters

\begin{document}\sloppy\sloppypar
\title{Combining information from photometric and spectroscopic data:\\
  Don't waste your time in spectrophotometric calibration!}
\shorttitle{don't waste your time in calibration!}
\author{
  BDJ\altaffilmark{\lick},
  DRW\altaffilmark{\lick},
  DFM\altaffilmark{\ccpp},
  DWH\altaffilmark{\ccpp, \cds, \mpia},
  CC\altaffilmark{\lick},
  others}
\shortauthors{bdj et al}
\altaffiltext{\lick}{University of California Lick Observatory}
\altaffiltext{\ccpp}{Center for Cosmology and Particle Physics, Department of Physics, New York University}
\altaffiltext{\cds}{Center for Data Science, New York University}
\altaffiltext{\mpia}{Max-Planck-Institut f\"ur Astronomie}

@@

If you don't believe the reported data noise variances $\sigma_n^2$,
it often makes sense to add in quadrature a ``jitter'' variance $s^2$.
This is a trivial modification of the chi-squared log-likelihood:
\begin{eqnarray}\label{eq:photometricLF}
\ln p(d\given\theta,s^2) &=& -\frac{1}{2}\,\sum_{n=1}^N
                             \left[\frac{[d_n -
                             \mu_n(\theta)]^2}{[\sigma_n^2 + (s\, \mu_n(\theta))^2]} + \ln(2\pi\,[\sigma_n^2 + s^2]) \right]
\quad .
\end{eqnarray}
This modification shows the necessity of including the log-variance
term.
This---equation~(\ref{eq:photometricLF})---is in fact the
log-likelihood function we are going to use for the photometric
measurements in what follows.  
In this formulation, if $\theta$ are the parameters of interest, the
jitter variance $s^2$ is a nuisance noise-model parameter representing
additonal fractional uncertainty in the flux.  The
Gaussian Process will be a generalization of this modification.

@@

\section{Experiments and Results}

@@

The separate treatment of mock and real data allows us to address two
separate issues.  The inference from mock data (with a realistic
calibration vector applied) provides a test of the methodology and the
effect of calibration uncertainties on the inferred parameters in the
case of a perfect mean model $mu(\theta)$.  The inference from real
data tests a combination of our methodology and the ability of the
population synthesis models to recover parameters inferred from
resolved star CMDs (i.e. some estimate of ground truth)

@@

\begin{figure}[h!]
%\includegraphics[width = 0.5 \textwidth]{figures/dfig_b192-g242_020.png}
\caption{The calibrated spectrum for \excluster, along with the
uncertainties (bottom panel).The right panel shows the
spectrophotometric calibration vector determined by
\citet{schiavon05}. \label{fig:data}}
\end{figure}

\subsection{List of Tests}

@@

\item Mock data: inference using photometry only, full spectroscopic model

@@

\subsection{Test Data}

The data we use to test our method consist of optical spectroscopy of
of galactic globular clusters from \citet{schiavon05} for which
optical and NIR photometry are availbale. The optical spectra were
obtained by drift-scanning a long-slit spectrograph across the
cluster.  The wavelength range of the spectra is $3500-6500$\AA. The
instrumental resolution is $\sim 3.1$\AA~ at $\sim 4500\AA$ and varies
with wavelength.  These spectra have been placed on the rest-frame,
though in what follows we will model the effect of redshift
uncertainties on the data.

An example spectrum (and the inferred spectrophotometric calibration
vector) is shown in Figure \ref{fig:data}.  As stated by
\citet{schiavon05}, the spectrophotometric calibration of these data are
only approximate.  When fitting the spectra we mask the region around
the bright OI5577\AA~ sky line, and we fit only to the spectrum with $3550\AA
< \lambda < 6500$\AA~ which are covered by the MILES spectral library
(see \S\ref{sec:cluster_model}).  Absorption lines of the interstellar
medium (ISM) can significantly affect the spectrum, so we
additionally mask the Calcium H and K lines and the NaD doublet.

Words about photometry

@@

\subsection{Mock Data}
Our mock data are designed to mimic the actual data as much as
possible.  Mock spectra and photometry are generated from the model
below for a given set of parameters $\theta^\prime$.  We then convolve
the spectrum with an instrumenal LSF, interpolate onto the observed
wavelength scale and add noise according to the
SNR of the real data.  For the photometry we assume a SNR of 10.  We
also mask the mock spectra at the location of significant sky lines
and ISM lines.

The mock spectrum is shown in Figure \ref{fig: mock spectrum}.  We
will refer to this spectrum as the \emph{perfectly calibrated}
mock spectrum.  We also produce a mock spectrum by dividing this
perfectly calibrated spectrum by the spectrophotometric calibration vector
taken from the real data
\citep[][Fig\ref{fig:ggc_spectrum}]{schiavon05}, which we refer to as
the \emph{uncalibrated} mock spectrum.

\subsection{Star Cluster Model}
We model the stellar cluster as a single age SSP. The spectrum of a
star cluster under this assumption is given by
\begin{eqnarray}\displaystyle
F_{\theta^\prime}(\lambda_o) & = &
\frac{m_*}{4\pi d_L^2(1+z)} \, S(\lambda, t, Z , \beta) \, e^ {-k_\lambda\tau_V} \\
\nonumber \\ 
\lambda_o & = & (1+z)\,\lambda 
\\
\theta^\prime & \equiv & \left[ m_*, t, Z, \tau_V, z, d_L, \beta \right]
\quad ,
\end{eqnarray} where the
parameters $\theta^\prime$ include 
the stellar mass $m_*$, 
effective optical depth to dust in the $V$-band $\tau_V$, 
luminosity distance $d_L$,
and redshift $z$.
The shape of the dust attenuation curve is given by $k_\lambda \equiv
\tau_\lambda/\tau_V$.  
The observed frame wavelengths $\lambda_o$ are related to the
restframe wavelengths $\lambda$ in the usual way.

The spectrum of one solar mass of stars that formed at lookback time
$t$ with metallicity $Z$ is given by $S(\lambda, t, Z [, \beta])$.
Here the additional parameters $\beta$ may include details of the
population synthesis model, such as the IMF slope and adjustments to
the stellar spectra or stellar evolutionary tracks. We will assume
$\beta$ to be fixed for the purposes of this exercise, with no loss of
generality but significant savings in computational time.  We will
also fix the shape of the attenuation curve, $k_\lambda $, to the
Milky way average of \citet{CCM89} with $R_V=3.1$ and assume that
$d_L$ is known to high precision from independent methods and can
therefore be treated as fixed.  In practice, $d_L$ is prefectly
degenerate with $m_*$. We calculate $S(\lambda, t, Z)$ using the
Flexible Stellar Population Synthesis code \citep[FSPS][]{fsps}.  We
use the Padova 2007 isochrones \citep{girardi00, bertilli94, marigo07}
and the MILES stellar spectral library \citep{miles_I, miles_II,
miles_III}.

@@

The model spectra $F_{\theta^\prime}$ are smoothed by convolution with
the line-spread function of the spectrograph and are interpolated to
the wavelength points of the of the observed spectrum while taking
into account the redshift. For computational simplicity we assume that
the internal velocity dispersion of the cluster is negligible compared
to the instrumental broadening, though in principle it can and should
also be modeled. he instrumental resolution is given by the parameter
$\sigma$.  {\color{blue}Discuss intrinsic resolution of model spectra
(MILES), what to do in situations where instrumental resolution is
higher, lower or comparable (ouch) to intrinsic velocity broadening.}

@@

Our prior distributions are in all cases top-hat functions, described
in Table \ref{tbl:priors}, though more informative prior distributions
may be used.  Informative priors may be especially appropriate for the
calibration parameters $\alpha$ if additional information about the
instrument is available; or for the parameters of interest $\theta$,
for example in the case that there is knowledge of the metallicity
distribution of the population from which a given cluster is drawn.

@@

\begin{figure*}[h!]
%\includegraphics[width=\textwidth]{figures/sfig_b192-g242_225_cal.png}
\caption{Spectra resulting from draws from the posterior parameter
space conditioned on the calibrated \excluster~ spectrum and
photometry.  Left: For each of 5 draws from the posterior we show from
top to bottom, three components of the model: the mean cluster
spectrum, the calibration vector, and the mean prediction of $\Delta$
for the Gaussian process.  Right: From top to bottom, the full model
({\it magenta}) compared to the observed, calibrated spectrum ({\it
grey}), the ratio of the observed to the modeled spectrum, and the
difference between the observed spectrum and the modeled spectrum
normalized by the uncertainty.
\label{fig:inferred_spectrum}}
\end{figure*}

\subsection{Results}

We first infer the parameters of interest $\theta$ for a calibrated
spectrum of one of the clusters, \excluster, using the full YY-band
photometry. In Figure \ref{fig:inferred_spectrum} we show the spectra
that result from draws of the parameters $[\theta,\alpha]$ from the
posterior distribution, as well as several of the individual
components of the model constructed from each draw. Figure
\ref{fig:inferred_spectrum} also shows the calibrated spectrum on
which the priors were conditioned, and residuals between the posterior
models and the observed data.  The agreement is quite good, though
some lines are not accurately predicted.  For example, the observed
Mg$b$ absorption appears stronger than the modeled
absorption.

@@

\subsection{Dependence on Photometric Data}
In this section we examine how the constraints on physical (and
calibration) parameters vary as a function of the number and type of
photometric points available. {\color{red} BDJ DO THIS TEST.}

@@

{\color{blue} BDJ BDJ discuss the relation of this technique to the typical use of
continuum normalization when fitting spectroscopy.}

@@

{\color{blue} BDJ - This paragraph is probably too calibration
focused. We want to focus more on spec + phot combinations.  At least,
this paragraph should not be in the discussion section.}
Fundamentally, calibration involves observing a `known' or `truth'
with an instrument, and using that observation to infer the effects of
the instrument.  However, we do not have knowledge of a single truth
for any object, only relative probabilities for what might be true.
In this sense, one should never use a single point estimate of the
calibration, the full distribution of the possible calibrations should
be inferred from our probabilistic knowledge of the calibration object
and forward propogated \citep[e.g.][]{lee11}. Once this is done, the
only distinction between 'calibration' objects and targets of interest
is the complexity of the model for the object and the level of
uncertainty in the calibration object spectrum.  If one is interested
in inferring parameters of interest of spectral models of
astrophysical objects, then there is often no reason not to use those
objects themselves as calibration objects.

BDJ misc issues, collected here in case we want to address them and to
keep me from cluttering up the main text: Model resolution and
instrument resolution.  How to interpret GP residuals, as a fn of S/N?
Photometric upper limits? Infer the wavelength solution as well -
stellar population (and telluric lines) as your arc. Combinations of
broadening in velocity space (physical) and in wavelength space
(instrumental)

>>>>

\begin{document}\sloppy\sloppypar
\title{Combining information from photometric and spectroscopic data:\\
  propogating uncertainties in spectrophotometric calibration}
\shorttitle{Combining Spectroscopy and Photometry}
\author{
  BDJ\altaffilmark{\lick},
  DRW\altaffilmark{\lick},
  DFM\altaffilmark{\ccpp},
  DWH\altaffilmark{\ccpp, \cds, \mpia},
  CC\altaffilmark{\lick},
  others}
\shortauthors{bdj et al}
\altaffiltext{\lick}{University of California Lick Observatory}
\altaffiltext{\ccpp}{Center for Cosmology and Particle Physics, Department of Physics, New York University}
\altaffiltext{\cds}{Center for Data Science, New York University}
\altaffiltext{\mpia}{Max-Planck-Institut f\"ur Astronomie}

@@

If you don't believe the reported data noise variances $\sigma_n^2$,
it often makes sense to add in quadrature a ``jitter'' variance
$s^2$.  For photometric data this ``jitter'' is often best described
as a fractional error, $(s\mu(\theta))^2$, corresponding for example
to an additional magnitude zero-point uncertainty.
This is a simple modification of the chi-squared log-likelihood:
\begin{eqnarray}\label{eq:photometricLF}
\ln p(d\given\theta,s^2) &=& -\frac{1}{2}\,\sum_{n=1}^N
                             \left[\frac{[d_n - \mu_n(\theta)]^2}
                             {[\sigma_n^2 + (s\,\mu_n(\theta))^2]} + 
                             \ln(2\pi\,[\sigma_n^2 + (s\,\mu(\theta))^2]) \right]
\quad .
\end{eqnarray}
This modification shows the necessity of including the log-variance
term.
This---equation~(\ref{eq:photometricLF})---is in fact the
log-likelihood function we are going to use for the photometric
measurements in what follows.  
In this formulation, if $\theta$ are the parameters of interest, the
jitter variance $s^2$ is a nuisance noise-model parameter representing
additonal fractional uncertainty in the flux.  The
Gaussian Process will be a generalization of this modification.

@@

\section{Data and Experiments}

@@

This separate treatment of mock and real data allows us to address two
separate issues.  The inference from mock data (with a realistic
calibration vector applied) provides a test of the methodology and the
effect of calibration uncertainties on the inferred parameters in the
case of a perfect mean model $mu(\theta)$.  The inference from real
data tests a combination of our methodology and the ability of the
population synthesis models to recover parameters inferred from
resolved star CMDs (i.e. some estimate of ground truth).

@@

\begin{figure}[h!]
%\includegraphics[width = 0.5 \textwidth]{figures/dfig_b192-g242_020.png}
\caption{The calibrated spectrum for \excluster, along with the
uncertainties (bottom panel).The right panel shows the
spectrophotometric calibration vector determined by
\citet{schiavon05}. \label{fig:ggc_data}}
\end{figure}

\subsection{List of Experiments}

@@

\item Mock data: inference using photometry only

@@

\subsection{Real Data}

The data we use to test our method consist of optical spectroscopy of
of galactic globular clusters from \citet{schiavon05} for which
optical and NIR photometry are available. The optical spectra were
obtained by drift-scanning a long-slit spectrograph across the
cluster.  The wavelength range of the spectra is $3500-6500$\AA. The
instrumental resolution is $\sim 3.1$\AA~ at $\sim 4500\AA$ and varies
with wavelength.  These spectra have been placed on the rest-frame,
though in what follows we will model the effect of redshift
uncertainties on the data.

An example spectrum (and the inferred spectrophotometric calibration
vector) is shown in Figure \ref{fig:ggc_data}.  As stated by
\citet{schiavon05}, the spectrophotometric calibration of these data are
only approximate.  When fitting the spectra we mask the region around
the bright OI5577\AA~ sky line, and we fit only to the spectrum with $3550\AA
< \lambda < 6500$\AA~ which are covered by the MILES spectral library
(see \S\ref{sec:cluster_model}).  Absorption lines of the interstellar
medium (ISM) can significantly affect the spectrum, so we
additionally mask the Calcium H and K lines and the NaD doublet.

Words about photometry.

@@

\subsection{Mock Data}
Our mock data are designed to mimic the actual data as much as
possible.  Mock spectra and photometry are generated from the model
described in below (\S\ref{sec:cluster_model}, equation
\ref{eq:MeanModel}) for a given set of parameters $\theta^\prime$.  We
then convolve the spectrum with an instrumenal LSF, interpolate onto
the observed wavelength scale and add noise according to the SNR of
the real data.  For the photometry we assume a SNR of 10.  We also
mask the mock spectra at the location of significant sky lines and ISM
lines that are present in the real data.

We will refer to these spectra as the \emph{perfectly calibrated} mock
spectrum.  An example is shown in Figure \ref{fig: mock data}. We also
produce mock spectra by dividing the perfectly calibrated spectra by
the spectrophotometric calibration vector taken from the real data
\citep[][Fig\ref{fig:ggc_spectrum}]{schiavon05}.  We refer to these as
the \emph{uncalibrated} mock spectra.

\begin{figure}[h!]
%\includegraphics[width = 0.5 \textwidth]{figures/dfig_b192-g242_020.png}
\caption{A mock spectrum for a star cluster with $10^5 M_\odot$ total
  stellar mass, an age of  9 Gyr, solar
  metallicity, and $A_V=0.5$ mag. \label{fig:mock_data}}
\end{figure}

\subsection{Star Cluster Model}
\label{sec:cluster_model}
We model the stellar cluster as a single age SSP. The spectrum of a
star cluster under this assumption is given by
\begin{eqnarray}\displaystyle
F_{\theta^\prime}(\lambda_o) & = &
\frac{m_*}{4\pi d_L^2(1+z)} \, S(\lambda, t, Z , \beta) \, e^ {-k_\lambda\tau_V} \\
\nonumber \\ 
\lambda_o & = & (1+z)\,\lambda 
\\
\theta^\prime & \equiv & \left[ m_*, t, Z, \tau_V, z, d_L, \beta \right]
\quad ,
\end{eqnarray} where the
parameters $\theta^\prime$ include 
the stellar mass $m_*$, 
effective optical depth to dust in the $V$-band $\tau_V$, 
luminosity distance $d_L$,
and redshift $z$.
The shape of the dust attenuation curve is given by $k_\lambda \equiv
\tau_\lambda/\tau_V$.  
The observed frame wavelengths $\lambda_o$ are related to the
restframe wavelengths $\lambda$ in the usual way.

The spectrum of one solar mass of stars that formed at lookback time
$t$ with metallicity $Z$ is given by $S(\lambda, t, Z [, \beta])$.
Here the additional parameters $\beta$ may include details of the
population synthesis model, such as the IMF slope and adjustments to
the stellar spectra or stellar evolutionary tracks. We will assume
$\beta$ to be fixed for the purposes of this exercise, with no loss of
generality but significant savings in computational time.  We will
also fix the shape of the attenuation curve, $k_\lambda $, to the
Milky way average of \citet{CCM89} with $R_V=3.1$ and assume that
$d_L$ is known to high precision from independent methods and can
therefore be treated as fixed.  In practice, $d_L$ is perfectly
degenerate with $m_*$. We calculate $S(\lambda, t, Z)$ using the
Flexible Stellar Population Synthesis code \citep[FSPS][]{fsps}.  We
use the Padova 2007 isochrones \citep{girardi00, bertilli94, marigo07}
and the MILES stellar spectral library \citep{miles_I, miles_II,
miles_III}.

@@

The model spectra $F_{\theta^\prime}$ are smoothed by convolution with
the line-spread function of the spectrograph and are interpolated to
the wavelength points of the of the observed spectrum while taking
into account the redshift. For computational simplicity we assume that
the internal velocity dispersion of the cluster is negligible compared
to the instrumental broadening, though in principle it can and should
also be modeled. The instrumental resolution is given by the parameter
$\sigma$.  {\color{blue}Discuss intrinsic resolution of model spectra
(MILES), what to do in situations where instrumental resolution is
higher, lower or comparable (ouch) to intrinsic velocity broadening.}

@@

Our prior distributions are in all cases top-hat functions, described
in Table \ref{tbl:priors}, though more informative prior distributions
may be used.  Informative priors may be especially appropriate for the
calibration parameters $\alpha$ if additional information about the
instrument is available; or for the parameters of interest $\theta$,
for example if there is knowledge of the metallicity
distribution of the population from which a given cluster is drawn.

@@

\begin{figure*}[h!]
%\includegraphics[width=\textwidth]{figures/sfig_b192-g242_225_cal.png}
\caption{Spectra resulting from draws from the posterior parameter
space conditioned on the calibrated \excluster~ spectrum and
photometry.  Left: For each of 5 draws from the posterior we show from
top to bottom, three components of the model: the mean cluster
spectrum, the calibration vector, and the mean prediction of $\Delta$
for the Gaussian process.  Right: From top to bottom, the full model
({\it magenta}) compared to the observed, calibrated spectrum ({\it
grey}), the ratio of the observed to the modeled spectrum, and the
difference between the observed spectrum and the modeled spectrum
normalized by the uncertainty.
\label{fig:speconly_calibrated}}
\end{figure*}

\section{Results}

For each test we want to show
\begin{itemize}

\item Posterior parameter pdfs (and joint pdfs).

\item Samples of the posterior prediction for the observations, and
  residuals thereform.

\item The calibration vector, and posterior samples of it.

\item Posterior samples of the intrinsic spectrum
\end{itemize}

\subsection{Mock data}

In Figure \ref{fig:speconly_calibrated} we show the result of our
inference from the spectrum and a single photometric measurement (SDSS
$g$) for a perfectly calibrated mock spectrum. To demonstrate the
quality of the fit we show both the mock
input spectrum and predictions of the observed spectra that result
from several draws of the model parameters
$[\theta, \alpha]$ from posterior distribution.  We also show samples
of the inferred calibration vector,
\begin{eqnarray} 
\frac{d}{\mu(\theta)} & = & e^{f(\alpha) + \tilde{\Delta}}
\quad ,
\end{eqnarray}
where $\tilde{\Delta}$ is the Gaussian process mean prediction for the
logarithmic residual between the data and the model.  For the
perfectly calibrated spectrum the true calibration vector (shown in
black) is one everywhere.  The spread of the inferred calibration
vectors approximately indicates our uncertainty in the spectrophotometric
calibration when no additional calibration information is available
and only a single photometric point is available.

Even in the case of a single photometric data point, several of the
physical parameters are well-constrained and only minimally biased,
e.g. age and metallicity.  These are the parameters that are
constrained well by the spectrum itself and are not strongly sensitive
to the overall shape of the spectral continuum.  Such parameters can
be reasonably well recovered by fitting continuum normalized spectra,
though hewre we have propogated the uncertainty in this normalization
into the uncertainties on the physical parameters.

On the other hand, the posterior PDF for the dust attenuation optical
depth $A_v$ is extremely broad.

The uncertainty on the total stellar mass incorporates both the
uncertainty in the photometric flux and the uncertainty in the $g$-
band mass-to-light ratio.

In Figure \ref{fig:speconly_uncalibrated} we show the results of
inference on an \emph{uncalibrated} mock spectrum.

\subsection{Real Data: NGCXXX} 
We first infer the parameters of interest $\theta$ for a calibrated
spectrum of one of the clusters, \excluster, using the full YY-band
photometry. In Figure \ref{fig:inferred_spectrum} we show the spectra
that result from draws of the parameters $[\theta,\alpha]$ from the
posterior distribution, as well as several of the individual
components of the model constructed from each draw. Figure
\ref{fig:inferred_spectrum} also shows the calibrated spectrum on
which the priors were conditioned, and residuals between the posterior
models and the observed data.

@@

@@

{\color{blue} BDJ discuss the relation of this technique to the typical use of
continuum normalization when fitting spectroscopy.}

@@

{\color{blue} BDJ - This paragraph is probably too calibration
focused. We want to focus more on spec + phot combinations.  At least,
this paragraph should not be in the discussion section.}
Fundamentally, calibration involves observing a `known' or `truth'
with an instrument, and using that observation to infer the effects of
the instrument.  However, we do not have knowledge of a single truth
for any object, only relative probabilities for what might be true.
In this sense, one should never use a single point estimate of the
spectrophotometric calibration, the full distribution of the possible
calibrations should be inferred from our probabilistic knowledge of
the calibration object and forward propogated
\citep[e.g.][]{lee11}. Indeed, even if the calibration object is known
perfectly, the observation of it will be noisy and thus so will the
inferred calibration. Once this forward propogation of calibration
uncertainty is done, the only distinction between 'calibration'
objects and targets of interest is the uncertainty in the model for
the object and the level of uncertainty in the observed object
spectrum -- this is why calibration objects are typically bright and
simple.  If one is interested in inferring parameters of interest of
spectral models of astrophysical objects, then there is often no
reason not to use those objects themselves as calibration objects
(especially if you expect your calibration to be spatially or
temporally variable, or to change with the brightness of the object)

BDJ misc issues, collected here in case we want to address them and to
keep me from cluttering up the main text: Model resolution and
instrument resolution, ombinations of broadening in velocity space
(physical) and in wavelength space (instrumental).  Photometric upper
limits? Infer the wavelength solution as well - stellar population
(and telluric lines) as your arc.
